### 2015-02-26 paper/exprs.tex
title, abstract, first round of corrections
second round of changes, acknowledgements

R-values in Patina are expressions that can evaluate to a layout.
Effectively, they are all the ways to construct values that can be deconstructed by L-values.
This layer of Patina involves the borrow checker, which ensures the borrowed reference guarantees.
In order to focus on the details of the borrow checker, we will pare down our types.
Variants, tuples, and recursive types have straightforward behavior here 
and would unnecessarily clutter the discussion.
Going forward, we will use this subset of our earlier language:

@@

\subsubsection*{TODO Move this into a foot note or something}
Rust itself and earlier models of Patina used a dramatically different form
of loan tracking and borrow checking. A universal set of loans it kept and threaded
through the type checker. Each loan generates a set of paths that cannot be accessed
without violating a loan guarantee. The borrow checker would then make sure that
no path used was in that set of forbidden paths.

This structure does not lend itself to proofs, so we devised this alternate system.
The system we present below is syntax driven and, we think, easier to understand.
Since proving this system matches the original would run into the problem we tried
to avoid in the first place, we satisfied ourselves with testing the two implementations.
The original model and this model agree on all the tests used in the original model.

\subsection*{Tracking Loans}

When we use a L-value in an expression, we must ensure the operation we are performing on it
does not violate the promises of existing loans, and we must ensure that the value it refers
to is fully initialized. When we create a borrowed reference, we must ensure that the
promises of the borrow can actually be upheld for the duration of the borrow.
In order to make these checks, we need to know what is borrowed, how it is borrowed, and for
how long it is borrowed. We will accomplish this by adding a stack of loans to every
level of the shadow heap:

@@

\subsection*{Shadow Typing with Banks}
Since our shadows now include loan information, we need to modify our shadow typing.
While we have include lifetimes in our model from the beginning, this is the first time
we will actually utilize them. Before we can discuss how to check banks, we must first
discuss what lifetimes are, where they come from, and what they give us.

@@

\subsubsection*{Bank Coherence}
As mentioned before, lifetimes come from scope blocks.
This means that every variable has a lifetime, which comes from the block it was defined in.
Since variables are deallocated at the end of the block in which they are defined,
that lifetime is a accurate model. We record the lifetimes of variables in a new context:

@@

\begin{mathpar}
  \infer[HT-UNINT]{ }{\tcb{\ltr}{\uninit}{\tyint}{mq}{\lt}} \and
  \infer[HT-UNOWN]{ }{\tcb{\ltr}{\uninit}{\own{\gt}}{mq}{\lt}} \and
  \infer[HT-UNREF]{(\lt,\lt')\in\ltr}{\tcb{\ltr}{\uninit}{\tyref{\lt'}{q}{\gt}}{mq}{\lt}} \and
  \infer[HT-INT]{ }{\tcb{\ltr}{\tyint}{\tyint}{mq}{\lt}} \and
  \infer[HT-OWN]{\tcb{\ltr}{\gs}{\gt}{mq}{\lt}}{\tcb{\ltr}{\own{\gs}}{\own{\gt}}{mq}{\lt}} \and
  %\infer[HT-REF]
    %{(\lt, \lt') \in \ltr \\ \tcb{\ltr}{\gs}{\gt}{mq}{\lt} \\ \gs\ \textrm{init}}
    %{\tcb{\ltr}{\refval{q}{\gs}}{\tyref{\lt'}{q}{\gt}}{mq}{\lt}}

  \and
  \infer[HT-REFMUT]
    {(\lt, \lt') \in \ltr \\ \tcb{\ltr}{\gs}{\gt}{mq}{\lt} \\ \gs\ \textrm{init}}
    {\tcb{\ltr}{\refval{\qmut}{\gs}}{\tyref{\lt'}{\qmut}{\gt}}{mq}{\lt}}
  \and
  \infer[HT-REFIMM]
    {(\lt, \lt') \in \ltr \\ \tcb{\ltr}{\gs}{\gt}{\qimm}{\lt} \\ \gs\ \textrm{init}}
    {\tcb{\ltr}{\refval{\qimm}{\gs}}{\tyref{\lt'}{\qimm}{\gt}}{mq}{\lt}}

@@

where $\gs\ \textrm{init}$ is a judgment that ensures a shadow is fully initialized,
i.e.\ it cannot reach a $\uninit$.
Note that the interior of immutable borrowed references is always controlled
by an immutable loan. This prevents mutable loans of the interior of immutable borrows,
which are nonsensical.

%need to think about how the qualifier of a borrowed reference impacts the kind of controlling loan of its contents. Ex: a mutable loan of the dereference of an immutable borrow should be forbidden. Dereferences of mutable borrows can be loaned either way...
%Perhaps the interior of an immutable borrow should be controlled by $\qimm$ always,
%and the interior of a mutable borrow should be controlled by the incoming controller.
%That seems sensible since immutable loans overpower mutable loans ... it's a similar pattern.

@@

Additionally, we can only use a path that is fully initialized, or our program might
try to touch a uninitialized value and get stuck.
This means that $\shallow{\gU}{\lv{x}{p}}{\gs}$ holds and that $\gs\ \textrm{init}$ holds. 
That is, enough memory is initialized to evaluate the L-value to a memory location,
and everything reachable from that location is initialized.
We use these dual judgments as a base to define our validity 
checks for reading, writing, and moving.
Reading will be more restricted than simple full initialization.
Writing will be more restricted than reading.
Moving will be more restricted than writing.

@@

\subsection*{Expression Evaluation}
Evaluating expressions is straightforward.
The only unusual aspect is the case of using the value at a L-value.
To prevent copies of affine values, we must overwrite the only value
with the used version. If it is a copyable value, then it will remain unchanged.
If it is an affine value, then it will be replaced with $\void$.
Note that this means the heap resulting from expression evaluation is either
unchanged or a more deinitialized version of the starting heap.
Also note that the result heap may temporarily contain an orphan when considered alone,
but not when considered together with the result layout.
\newline

>>>>

R-values in Patina are expressions that can evaluate to a layout.
This layer of Patina involves the borrow checker, which ensures the borrowed reference guarantees.
In order to focus on the details of the borrow checker, we will pare down our types.
Variants, tuples, and recursive types have straightforward behavior here 
and would unnecessarily clutter the discussion.
Going forward, we will use this subset of our earlier language:

@@

\subsection*{Tracking Loans\footnotemark}

\footnotetext {
Rust itself and earlier models of Patina used a dramatically different form
of loan tracking and borrow checking. A universal set of loans is kept and threaded
through the type checker. Each loan generates a set of paths that cannot be accessed
without violating a loan guarantee. The borrow checker would then make sure that
no path used was in that set of forbidden paths.

This structure does not lend itself to proofs, so we devised this alternate system.
The system we present below is syntax driven and, we think, easier to understand.
Since proving this system matches the original would run into the problem we tried
to avoid in the first place, we satisfied ourselves with testing the two implementations.
The original model and this model agree on all the tests used in the original model.
}

When we use an L-value in an expression, we must ensure the operation we are performing on it
does not violate the promises of existing loans, and we must ensure that the value it refers
to is fully initialized. When we create a borrowed reference, we must ensure that the
promises of the borrow can actually be upheld for the duration of the borrow.
In order to make these checks, we need to know what is borrowed, how it is borrowed, and for
how long it is borrowed. We will accomplish this by adding a stack of loans to every
level of the shadow heap:

@@

\subsection*{Shadow Typing with Banks}
Since our shadows now include loan information, we need to modify our shadow typing.
While we have included lifetimes in our model from the beginning, this is the first time
we will actually utilize them. Before we can discuss how to check banks, we must first
discuss what lifetimes are, where they come from, and what they give us.

@@

\subsubsection*{Bank Coherence}
As mentioned before, lifetimes come from scope blocks.
This means that every variable has a lifetime, which comes from the block it was defined in.
Since variables are deallocated at the end of the block in which they are defined,
that lifetime is an accurate model. We record the lifetimes of variables in a new context:

@@

\begin{mathpar}
  \infer[HT-UNINT]{ }{\tcb{\ltr}{\uninit}{\tyint}{mq}{\lt}} \and
  \infer[HT-UNOWN]{ }{\tcb{\ltr}{\uninit}{\own{\gt}}{mq}{\lt}} \and
  \infer[HT-UNREF]{(\lt,\lt')\in\ltr}{\tcb{\ltr}{\uninit}{\tyref{\lt'}{q}{\gt}}{mq}{\lt}} \and
  \infer[HT-INT]{ }{\tcb{\ltr}{\tyint}{\tyint}{mq}{\lt}} \and
  \infer[HT-OWN]{\tcb{\ltr}{\gs}{\gt}{mq}{\lt}}{\tcb{\ltr}{\own{\gs}}{\own{\gt}}{mq}{\lt}} \and
  \and
  \infer[HT-REFMUT]
    {(\lt, \lt') \in \ltr \\ \tcb{\ltr}{\gs}{\gt}{mq}{\lt} \\ \gs\ \textrm{init}}
    {\tcb{\ltr}{\refval{\qmut}{\gs}}{\tyref{\lt'}{\qmut}{\gt}}{mq}{\lt}}
  \and
  \infer[HT-REFIMM]
    {(\lt, \lt') \in \ltr \\ \tcb{\ltr}{\gs}{\gt}{\qimm}{\lt} \\ \gs\ \textrm{init}}
    {\tcb{\ltr}{\refval{\qimm}{\gs}}{\tyref{\lt'}{\qimm}{\gt}}{mq}{\lt}}

@@

where ``$\gs\ \textrm{init}$'' in HT-REFMUT and HT-REFIMM is a judgment 
that ensures a shadow is fully initialized,
i.e.\ it cannot reach a $\uninit$.
Note that the interior of immutable borrowed references is always controlled
by an immutable loan. This prevents mutable loans of the interior of immutable borrows,
which are nonsensical.

@@

Additionally, we can only use a path that is fully initialized, or our program might
try to touch an uninitialized value and get stuck.
This means that $\shallow{\gU}{\lv{x}{p}}{\gs}$ holds and that $\gs\ \textrm{init}$ holds. 
That is, enough memory is initialized to evaluate the L-value to a memory location,
and everything reachable from that location is initialized.
We use these dual judgments as a base to define our validity 
checks for reading, writing, and moving.
Reading will be more restricted than simple full initialization.
Writing will be more restricted than reading.
Moving will be more restricted than writing.

@@

\subsection*{Expression Evaluation}
Evaluating expressions is straightforward.
The only unusual aspect is the case of using the value at a L-value.
To prevent copies of affine values, we must overwrite the only value
with the used version. If it is a copyable value, then it will remain unchanged.
If it is an affine value, then it will be replaced with $\void$.
Note that this means the heap resulting from expression evaluation is either
unchanged or a more deinitialized version of the starting heap.
Also note that the result heap may temporarily contain a leak when considered alone,
but not when considered together with the result layout.
\newline

### 2015-02-26 paper/paths.tex
second round of changes, acknowledgements

\subsection*{Layout and Cell Typing}
Typing for layouts and cells is what you would expect.
However, we restrict the type of $\void$ to the possible types for other
kinds of cells, i.e. $\tyint$, unique pointers, and borrowed references.
This means that multi-cell layout types (variants, tuples, and recursive types) imply
a non-$\void$ value.
\newline

@@

\subsection*{Heap Well-Formedness}
With runtime typing now in hand, we can say what it means for the heap to be well-formed.
We say a heap $H$ is described by a heap type $\gS$ and a discriminant context $\gY$ if
every allocation in $H$ has the type assigned to it by $\gS$
and $\gY$ correctly records the discrimiants of all the variants in $H$.
Formally,

\begin{mathpar}
\mprset{flushleft}
\infer
{\dom{H} = \dom{\gS} \\\\
\forall \ga \in \dom(\gS).~\gS(\ga)~\textrm{closed} \\\\
\forall \ga \in \dom{H}.~\tc{\gS;\gY}{H(\ga)}{\gS(\ga)} \\\\
\forall (\ga,r) \in \dom{\gY}.~\Read{H}{\addr{\ga}{r}}{\varval{\gY(\ga,r)}{l}} \\\\
\forall \ga, \ga', \ga'' \in \dom H.\ \Read{H}{\ga}{\own{\ga''}} \land
				      \Read{H}{\ga'}{\own{\ga''}} \implies \ga = \ga' }
{\tc{}{H}{\gS;\gY}}
\end{mathpar}

@@

\section*{Shadow Coherence}
Of course, we also need to check that our shadow heap correctly models
the actual heap. For the most part, it is exactly the relation you would expect.
The use of \textsc{init} in \textsc{LS-REF} and \textsc{LS-VAR} ensures that
the pointees of borrowed references and the payload of initialized variants
are fully initialized.
The $\textsc{droppable}(l)$ judgment checks that $l$ owns no initialized heap memory,
i.e. $\own{\addr{\ga}{r}}$ is not reachable from $l$ except through references.
These layouts are safe to deallocate because we cannot leak heap memory by doing so.
The $\tc{V}{H}{\gU}$ judgment ensures that the shadow heap corresponds to the real heap.
\newline

@@

\section*{Shadow Preservation}
The second main preservation lemma for our core operations.
The initialization data, in the form of a shadow,
is preserved by path evaulation and \textsc{read}
so that the resulting layout is correctly described by the shadow.

@@

\section*{Path Progress}
Finally, we have all we need to proof path evaluation progress under the right circumstances.
Assuming all our contexts and the heap are well-formed,
then a shallowly initialized, well-typed path can evaluate to some route.
We can then rely to path preservation to show that route has the same type as the path.
Read safety then lets us show that we can read a layout at that route
and that it has the same type. Shadow preservation guarantees us that our
model of the layout reflects reality. All of the key pieces we need for
the higher layers of Patina.

\begin{lem}[Path Progress]
  If $\tc{}{H}{\gS;\gY}$, $\gG;V\vdash\gS$, $\tc{}{\gU}{\gG}$, $\tc{V}{H}{\gU}$,
  $\tc{\gG}{\lv{x}{p}}{\gt}$, and $\shallow{\gU}{\lv{x}{p}}{\gs}$
  then there is some $\addr{\ga}{r}$ such that $\ev{V;H}{\lv{x}{p}}{\addr{\ga}{r}}$.
\end{lem}

>>>>

\subsection*{Layout and Cell Typing}
Typing for layouts and cells is what you would expect.
However, we restrict the type of $\void$ to the possible types for other
kinds of cells, i.e. $\tyint$, unique pointers, and borrowed references.
This means that multi-cell layout types (variants, tuples, and recursive types) imply
a non-$\void$ value (though they can still contain $\void$s).
\newline

@@

\subsection*{Heap Well-Formedness}
With runtime typing now in hand, we can say what it means for the heap to be well-formed.
We say a heap $H$ is described by a heap type $\gS$ and a discriminant context $\gY$ if
every allocation in $H$ has the type assigned to it by $\gS$
and $\gY$ correctly records the discrimiants of all the variants in $H$.
The last hypothesis enforces the uniqueness of unique pointers by requiring
that if a unique pointer is at any address, then that address is unique.
This strengthens our induction hypothesis enough to show that no memory has
multiple owners, which we need to prove that no double frees occur.
Ensuring every piece of memory has at least one owner is handled elsewhere.
Formally,

\begin{mathpar}
\mprset{flushleft}
\infer
{\dom{H} = \dom{\gS} \\\\
\forall \ga \in \dom(\gS).~\gS(\ga)~\textrm{closed} \\\\
\forall \ga \in \dom{H}.~\tc{\gS;\gY}{H(\ga)}{\gS(\ga)} \\\\
\forall (\ga,r) \in \dom{\gY}.~\Read{H}{\addr{\ga}{r}}{\varval{\gY(\ga,r)}{l}} \\\\
\forall \ga, \ga', \ga'' \in \dom H.\ \forall r, r.\ 
    \Read{H}{\addr{\ga}{r}}{\own{\ga''}} \land
    \Read{H}{\addr{\ga'}{r'}}{\own{\ga''}} \implies \addr{\ga}{r} = \addr{\ga'}{r'} }
{\tc{}{H}{\gS;\gY}}
\end{mathpar}

@@

\section*{Shadow Checking the Heap}
Of course, we also need to check that our shadow heap correctly models
the actual heap. For the most part, it is exactly the relation you would expect.
The use of \textsc{init} in \textsc{LS-REF} and \textsc{LS-VAR} ensures that
the pointees of borrowed references and the payload of initialized variants
are fully initialized.
The $\textsc{droppable}(l)$ judgment checks that $l$ owns no initialized heap memory,
i.e. $\own{\addr{\ga}{r}}$ is not reachable from $l$ except through references.
These layouts are safe to deallocate because we cannot leak heap memory by doing so.
The $\tc{V}{H}{\gU}$ judgment ensures that the shadow heap corresponds to the real heap.
\newline

@@

\section*{Shadow Preservation}
The initialization data, in the form of a shadow,
is preserved by path evaulation and \textsc{read}
so that the resulting layout is correctly described by the shadow.

@@

\section*{Path Progress}
Finally, we have all we need to proof path evaluation progress under the right circumstances.
Assuming all our contexts and the heap are well-formed,
then a shallowly initialized, well-typed path can evaluate to some route.
We can then rely on path preservation to show that route has the same type as the path.
Read safety then lets us show that we can read a layout at that route
and that it has the same type. Shadow preservation guarantees us that our
model of the layout reflects reality. All of the key pieces we need for
the higher layers of Patina.

\begin{lem}[Path Progress]
  \raggedright{}
  If $\tc{}{H}{\gS;\gY}$, $\gG;V\vdash\gS$, $\tc{}{\gU}{\gG}$, $\tc{V}{H}{\gU}$,
  $\tc{\gG}{\lv{x}{p}}{\gt}$, and $\shallow{\gU}{\lv{x}{p}}{\gs}$
  then there is some $\addr{\ga}{r}$ such that $\ev{V;H}{\lv{x}{p}}{\addr{\ga}{r}}$.
\end{lem}
